GVN-Hoist: Hoisting Computations from Branches

### 2016-08-23 hoist.tex
add hoisting bug refs
cleanup finding candidates

A lot of bug reports in GCC and LLVM bugzillas, no papers on this topic yet.

@@

\subsection{}
The algorithm for code hoisting uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Control Flow Graph (CFG) and the Dominance (DOM) and Post-Dominance (PDOM)
  relations \cite{dragonbook};
\item Single Entry Single Exit (SESE) \cite{sese} and Single Entry Multiple Exit
  (SEME) regions;
\item Static Single Assignment (SSA) \cite{cytron};
\item Global Value Numbering (GVN) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use GVN.  Each expression is given a
  unique number and the expressions that the compiler can prove identical are
  given the same number;
\item Memory SSA \cite{novillo2007memory}: memory operations that the compiler
  is able to prove in dependence are linked through use-def chains;
\end{itemize}

For the code hoisting we follow a very systematic approach to writing a compiler
optimization pass. As such this pass can be divided into the following five
steps:

\subsection{Finding the candidates}
The first step is to find if a set of instructions perform identical
computations. Since our code hoisting algorithm is based on GVN, we need the
facility to compute GVN of instructions.  In llvm \cite{llvm}, GVN
infrastructure is already set up and works well for most kind of instructions
except loads and stores so we compute the GVN of loads and stores separately. In
order to value number loads, we value number (hash) the address from where the
value is to be loaded. For stores, we value number (hash) the address as well as
the value to be stored at that address.

The process of computing GVN can be on-demand (as we come across an instruction)
or, precomputed (computing GVN of all the instructions beforehand). Which
process to choose is determined by the scope of code-hoisting we want to
perform. In a pessimistic approach (Section-\ref{subsec:pessimistic}) where we
want to hoist limited set of instructions from the sibing branches as we iterate
the DFS tree bottom-up, it is sufficient to compute values on-demand. Whereas,
in the optimistic approach (Section-\ref{subsec:optimistic}), where we want to
hoist as many instructions as possible, would require values to be precomputed.

Whatever approach we take, to find a candidate all we need to do is to compare
the value number of computations.

>>>>

A lot of bug reports in GCC and LLVM bugzillas
\cite{GCCCodeHoistingBugs,LLVMCodeHoistingBugs}, no papers on this topic yet.

@@

The algorithm for code hoisting uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Control Flow Graph (CFG) and the Dominance (DOM) and Post-Dominance (PDOM)
  relations \cite{dragonbook};
\item Single Entry Single Exit (SESE) \cite{sese} and Single Entry Multiple Exit
  (SEME) regions;
\item Static Single Assignment (SSA) \cite{cytron};
\item Global Value Numbering (GVN) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use GVN.  Each expression is given a
  unique number and the expressions that the compiler can prove identical are
  given the same number;
\item Memory SSA \cite{novillo2007memory}: memory operations that the compiler
  is able to prove in dependence are linked through use-def chains.
\end{itemize}

The code-hoisting pass can be divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}
\item find candidates to hoist and compute hoisting position,
\item check whether moving the code is legal and profitable,
\item transform the code.
\end{itemize}

\subsection{Finding the candidates}
The first step is to find a set of instructions that perform identical
computations: this is performed by a linear scan of all instructions of the
program and classifying all instructions by their value given by GVN.  The
current implementation of GVN in LLVM has some limitations when it comes to
loads and stores so we compute the GVN of loads and stores separately.  Our
short-time solution to value number loads is to hash the address from where the
value is to be loaded. For stores, we value number the address as well as the
value to be stored at that address.

Another limitation of the current GVN implementation in LLVM is that the
instructions dependent on the loads will not get numbered correctly, and so
after hoisting all candidates we need to rerun the GVN analysis in order to
discover new candidates now available after having hoisted load instructions.
This limitation should be addressed in a new implementation of the GVN based on
MemorySSA, that would better account for equivalent loads and their dependent
instructions.

The process of computing GVN can be on-demand (as we come across an instruction)
or, precomputed (computing GVN of all the instructions beforehand). Which
process to choose is determined by the scope of code-hoisting we want to
perform. In a pessimistic approach (Section-\ref{subsec:pessimistic}) where we
want to hoist limited set of instructions from the sibling branches as we iterate
the DFS tree bottom-up, it is sufficient to compute values on-demand. Whereas,
in the optimistic approach (Section-\ref{subsec:optimistic}), where we want to
hoist as many instructions as possible, would require values to be precomputed.

Once the instructions have been classified in equivalence classes, we compute
for each group of equivalent instructions a point in the program that is both
legal and profitable for the code-hoisting.
